Bayesian Inference in Numerical Cognition: A Tutorial Using JASP

### 2019-12-07 jncTutorial/manuscript.Rmd
minor edits throughout and start conclusion

Like any other discipline within the psychological and behavioral sciences, researchers in numerical cognition rely on statistical inference, and hypothesis testing in particular, to verify theoretical claims against observed data. In recent years, Bayesian inference has become increasingly popular as an alternative to classical null hypothesis significance testing (NHST). Bayesian inference operates under a fundamentally different framework from NHST; instead of hard "accept/reject" decisions about a null hypothesis, Bayesian inference works by quantifying knowledge about competing hypotheses in light of observed data. Hypotheses that do a better job of predicting data will receive increased support, whereas hypotheses that don't predict the data well will receive decreased support. Under this framework, researchers can mathematically quantify belief about parameters

@@

Even though the reader of this paper is probably familiar with the general concept of hypothesis testing, we believe that it is nonetheless instructive to go over the basics.

## Hypotheses
Hypothesis testing can be viewed as a process for validating general claims based on a small sample of the general population. For illustrative purposes, suppose that we are policy makers confronted with a new study programme which claims to increase the mathematical abilities of sixth graders in the United States. To ease the exposition we use a (fictitious) scale called the Scale for Advancing Mathematical Ability (SAMA) to measure mathematical ability. As policy makers we know that the population mean SAMA score is \( \mu_{0}=50 \) and that the standard deviation is \( \sigma=10 \) for the population of sixth graders that are taught using the standard materials in the US.

@@

To nonetheless infer the validity of the working hypothesis \( \mathcal{H}_{1} \), we run an experiment in which \( n=90 \) sixth graders are taught with the new study programme. Suppose that these particular participants scored an average SAMA score of \( \bar{x} = 52.07 \). Clearly, this sample mean is larger than the status quo population mean of \( 50 \), but can we assume that this finding generalises to the whole population? If the answer is "yes" and we declare the working hypothesis to be verified, then how confident would we be about such a decision? Below we recall how \( p \)-values are typically used to make hard "yes" and "no" decisions, and we remind the reader why a \( p \)-value does not quantify the (un)certainty of such decisions.

## Statistical models
To deduce how hypotheses, i.e., claims about the population such as \( \mu \), affect the data, such as \( \bar{x} \), we have to link the former to the latter with a distribution, for instance, with a normal distribution. The population parameters, the distribution that links them to the data, and the data together make up a (statistical) model, which allows for quantitative predictions. For instance, the claim that the treatment population is identitical to the current population of sixth graders implies the null model \( \mathcal{M}_{0} \) comprised of (1) the null hypothesis \( \mathcal{H}_{0}: \mu = 50 \), (2) the assumption \( \sigma = 10 \), and (3) normally distributed SAMA scores. When all three assumptions hold true, we then deduce that the \emph{potential} average SAMA score consisting of a group of \( n = 90 \) participants is expected to fall within the interval \( 47.93 = 50 - 1.96 \frac{10}{\sqrt{n}} \) and \( 52.07 = 50 + 1.96 \frac{10}{\sqrt{n}} \) with 95% chance. <!-- 
-->
\footnote{We use ``chance'' to refer to probabilistic statements about the data.}
<!-- 
-->

\subsection{Null hypothesis significance testing}
Reasoning from the general population under the assumption that the parameters are known to derive statements about potential outcomes forms the basis of null hypothesis signficance testing. By definition the \( p \)-value is the chance of seeing the observed outcome and more extreme --but \emph{unobserved}-- outcomes under the null.

@@

\subsection{Bayesian testing}
Bayesian testing addresses these problems as follows: Firstly, it compares the two hypotheses on equal footing by also operationalising the alternative model. For the case at hand, \( \mathcal{M}_{1} \) would be comprised of the assumptions (1') \( \mathcal{H}_{1}: \mu > 50 \), (2) \( \sigma =10 \) as before, and (3) the data being normally distributed as before. By doing so, violations of normality and the assumed standard deviation will penalize both models equally. Secondly, to reason from a particular observed outcome to the general population, it employs Bayes rule. To do so, we are required to express our uncertainty about the two hypotheses in terms of so-called prior model probabilities before data observation.

Prior model probabilities help us contextualise the testing problem. Working as policy makers, we might have seen many claims of improved study programmes that did not work out well. For instance, if 4-to-1 claims of improved study programmes led to null results, we might set \( P( \mathcal{M}_{1}) = 0.2 \) and \( P(\mathcal{M}_{0}) = 0.8 \). On the other hand, the designers of the new study programme to increase SAMA scores might choose \( P( \mathcal{M}_{1}) = 0.7 \) and \( P(\mathcal{M}_{0}) = 0.3 \) instead. If \( \mathcal{M}_{1} \) and \( \mathcal{M}_{0} \) are the only models under consideration we have \( 0 < P( \mathcal{M}_{0}) < 1 \) and \( P( \mathcal{M}_{1})= 1 - P( \mathcal{M}_{0}) \). The Bayes factor \( \textnormal{BF}_{10}(d) \) is a function of the observed data \( d \) that updates these prior model probabilities to posterior model probabilities, as follows:

@@

## Bayes factor interpretations
Prior model probabilities are context dependent, whereas Bayes factors only rely on the statistical design. To simplify notation we drop the reference to the data in the Bayes factor and denote them as \( \textnormal{BF}_{10} \), where the 1 in the subscript refers to the alternative model, and the 0 to the null model. Bayes factor ranges from zero to \( + \infty \), and \( \textnormal{BF}_{10} = 8 \) implies that the observed data are 8 times more likely under the alternative compared to the null model. For Bayes factors smaller than one, we typically take the reciprocal to make them easier to interpret. For instance, \( \textnormal{BF}_{10} = 0.2 \) is easier interpret as \( \textnormal{BF}_{01} = 1/\textnormal{BF}_{10} = 5 \), which means that the observed data are 5 times more likely under the null compared to the alternative. Bayes factors are best viewed as a continuous measure of evidence. In general, a value of \( \textnormal{BF}_{10} \) larger than one is perceived to provide evidence for the alternative against the null. Also, if  \( \textnormal{BF}_{10} \) grows to \( \infty \), the posterior model probability \( P(\mathcal{M}_{1} \mid d) \) grows to one, thus the more certain we will be about the adequacy of the alternative.

@@

\begin{tabular}{rl}
Bayes factor & Evidence\\
\hline
1-3 & anecdotal\\
3-10 & moderate\\
10-30 & strong\\
30-100 & very strong\\
\( > \)100 & extreme\\
\end{tabular}

@@

<!--@Tom, this is how I normally view this. Furthermore, in linear regression, we can discuss how the uncertainty of the testing problem and the uncertainty of the estimation are combined. -->
What is important to realize is that a test addresses a different question than an estimation. The Bayes factor focuses on the question "Does the effect exists?", whereas the posterior focuses on the question "If it exists, then how large is it?". This is why the (Bayesian) testing problem compares the model where the population effect is fixed at zero against one where it is free to vary. Once the data have convinced us that the population effect \( \delta \) varies freely would we be interested in the magnitude of the effect.

@@

We use an example inspired by @hollowayAnsari2009, who examined the relationship between basic numerical processing and higher mathematical skills in children. Specifically, they examined whether childrens' mathematical achievement scores was related to their ability to compare symbolic and nonsymbolic number magnitudes. In this example, we will recreate one specific component of their analysis; namely, we will use Bayesian linear regression to examine the extent to which mathematical fluency scores are predicted by childrens' standardized symbolic and nonsymbolic distance effects, calculated by subtracting mean RTs for large numerical distances from mean RTs for small numerical distances and then dividing by the mean RT for large distances. In their original paper, @hollowayAnsari2009 found that the childrens' symbolic distance effect was a significant predictor of fluency, $\beta = -0.236$, $p<0.05$, whereas the nonsymbolic distance effect did not predict fluency, $\beta=0.063$, $p>0.05$.

@@

To demonstrate a Bayesian factorial analysis of variance, we use a data set based on @campbellFugelsang2001, who investigated the effects of problem size and numerical surface form on response times in an addition verification task. In addition to the usual effects of problem size (small problems are solved faster than large problems) and problem format (digit problems are solved faster than word problems), they found an interaction between size and format, which they interpreted as support for an interactive model of mental arithmetic where factors that affect encoding processes (e.g., surface format) can directly impact calculation processes.

@@

>>>>

Like any other discipline within the psychological and behavioral sciences, researchers in numerical cognition rely on statistical inference, and hypothesis testing in particular, to verify theoretical claims against observed data. In recent years, Bayesian inference has become increasingly popular as an alternative to classical null hypothesis significance testing (NHST). Bayesian inference operates under a fundamentally different framework from NHST; instead of hard "accept/reject" decisions about a null hypothesis, Bayesian inference works by quantifying knowledge about competing hypotheses in light of observed data. Hypotheses that do a better job of predicting data will receive increased support, whereas hypotheses that don't predict the data well will receive decreased support.

@@

In this section, we will review and elaborate on the concept of hypothesis testing. Even though the reader of this paper is probably familiar with the general concept of hypothesis testing, we believe that it is nonetheless instructive to go over the basics.

## Hypotheses
Hypothesis testing can be viewed as a process for validating general claims based on a small sample of the general population. To illustrate, suppose that we are policy makers confronted with a new study programme which claims to increase the mathematical abilities of sixth graders in the United States. To ease the exposition we use a (fictitious) scale called the Scale for Advancing Mathematical Ability (SAMA) to measure mathematical ability. As policy makers we know that the population mean SAMA score is \( \mu_{0}=50 \) and that the standard deviation is \( \sigma=10 \) for the population of sixth graders that are taught using the standard materials in the US.

@@

Nonetheless, we can infer the validity of the working hypothesis \( \mathcal{H}_{1} \) by running an experiment in which \( n=90 \) sixth graders are taught with the new study programme. Suppose that these particular participants scored an average SAMA score of \( \bar{x} = 52.07 \). Clearly, this sample mean is larger than the status quo population mean of \( 50 \), but can we assume that this finding generalises to the whole population? If the answer is "yes" and we declare the working hypothesis to be verified, then how confident would we be about such a decision? Below we recall how \( p \)-values are typically used to make hard "yes" and "no" decisions, and we remind the reader why a \( p \)-value does not quantify the (un)certainty of such decisions.

## Statistical models
To deduce how hypotheses, i.e., claims about the population such as \( \mu \), affect the data, such as \( \bar{x} \), we have to link the former to the latter with a distribution. For instance, let us suppose a normal distribution provides this link. The population parameters, the distribution that links them to the data, and the data together make up a (statistical) model, which allows for quantitative predictions. For instance, the claim that the treatment population is identitical to the current population of sixth graders implies the null model \( \mathcal{M}_{0} \) comprised of (1) the null hypothesis \( \mathcal{H}_{0}: \mu = 50 \), (2) the assumption \( \sigma = 10 \), and (3) normally distributed SAMA scores. When all three assumptions hold true, we can mathematically deduce that the \emph{potential} average SAMA score consisting of a group of \( n = 90 \) participants is expected to fall within the interval \( 47.93 = 50 - 1.96 \frac{10}{\sqrt{n}} \) and \( 52.07 = 50 + 1.96 \frac{10}{\sqrt{n}} \) with 95% chance. <!-- 
-->
\footnote{We use ``chance'' to refer to probabilistic statements about the data.}
<!-- 
-->

\subsection{Null hypothesis significance testing}
Reasoning from the general population under the assumption that the parameters are known to derive statements about potential outcomes forms the basis of null hypothesis signficance testing. By definition, the \( p \)-value is the chance of seeing the observed outcome and more extreme --but \emph{unobserved}-- outcomes under the null.

@@

\subsection{Bayesian testing}
Bayesian testing addresses these problems as follows. First, it compares the two hypotheses on equal footing by also operationalising the alternative model. For the case at hand, \( \mathcal{M}_{1} \) would be comprised of the assumptions (1') \( \mathcal{H}_{1}: \mu > 50 \), (2) \( \sigma =10 \) as before, and (3) the data being normally distributed as before. By doing so, violations of normality and the assumed standard deviation will penalize both models equally. Secondly, to reason from a particular observed outcome to the general population, it employs Bayes rule. To do so, we are required to express our uncertainty about the two hypotheses in terms of so-called *prior model probabilities* before data observation.

Prior model probabilities help us contextualise the testing problem. Working as policy makers, we might have seen many claims of improved study programmes that did not work out well. For instance, past claims of improved study programmes led to null results four times out of five (i.e., we place 4-to-1 odds against the claim of improvement), we might set \( P( \mathcal{M}_{1}) = 0.2 \) and \( P(\mathcal{M}_{0}) = 0.8 \). On the other hand, the designers of the new study programme to increase SAMA scores might feel strongly about the benefits of the problem, and so might choose \( P( \mathcal{M}_{1}) = 0.7 \) and \( P(\mathcal{M}_{0}) = 0.3 \) instead. If \( \mathcal{M}_{1} \) and \( \mathcal{M}_{0} \) are the only models under consideration we have \( 0 < P( \mathcal{M}_{0}) < 1 \) and \( P( \mathcal{M}_{1})= 1 - P( \mathcal{M}_{0}) \). The Bayes factor \( \textnormal{BF}_{10}(d) \) is a function of the observed data \( d \) that updates these prior model probabilities to posterior model probabilities, as follows:

@@

## Bayes factor interpretations
Prior model probabilities are context dependent, whereas Bayes factors only rely on the statistical design. To simplify notation we drop the reference to the data in the Bayes factor and denote it as \( \textnormal{BF}_{10} \), where the 1 in the subscript refers to the alternative model, and the 0 to the null model. Bayes factors range from zero to \( + \infty \), and \( \textnormal{BF}_{10} = 8 \) implies that the observed data are 8 times more likely under the alternative compared to the null model. For Bayes factors smaller than one, we typically take the reciprocal to make them easier to interpret. For instance, \( \textnormal{BF}_{10} = 0.2 \) is easier interpret as \( \textnormal{BF}_{01} = 1/\textnormal{BF}_{10} = 5 \), which means that the observed data are 5 times more likely under the null compared to the alternative. Bayes factors are best viewed as a continuous measure of evidence. In general, a value of \( \textnormal{BF}_{10} \) larger than one is perceived to provide evidence for the alternative against the null. Also, if  \( \textnormal{BF}_{10} \) grows to \( \infty \), the posterior model probability \( P(\mathcal{M}_{1} \mid d) \) grows to one, thus the more certain we will be about the adequacy of the alternative.

@@

\begin{center}
\begin{tabular}{rl}
Bayes factor & Evidence\\
\hline
1-3 & anecdotal\\
3-10 & moderate\\
10-30 & strong\\
30-100 & very strong\\
\( > \)100 & extreme\\
\end{tabular}
\end{center}

@@

<!--@Tom, this is how I normally view this. Furthermore, in linear regression, we can discuss how the uncertainty of the testing problem and the uncertainty of the estimation are combined. -->
What is important to realize is that a test addresses a different question than an estimation. The Bayes factor focuses on the question "Does the effect exist?", whereas the posterior focuses on the question "If it exists, then how large is it?". This is why the (Bayesian) testing problem compares the model where the population effect is fixed at zero against one where it is free to vary. Once the data have convinced us that the population effect \( \delta \) varies freely, only then would we be interested in the magnitude of the effect.

@@

We use an example inspired by @hollowayAnsari2009, who examined the relationship between basic numerical processing and higher mathematical skills in children. Specifically, they examined whether childrens' mathematical achievement scores were related to their ability to compare symbolic and nonsymbolic number magnitudes. In this example, we will recreate one specific component of their analysis; namely, we will use Bayesian linear regression to examine the extent to which mathematical fluency scores are predicted by childrens' standardized symbolic and nonsymbolic distance effects, calculated by subtracting mean RTs for large numerical distances from mean RTs for small numerical distances and then dividing by the mean RT for large distances. In their original paper, @hollowayAnsari2009 found that the childrens' symbolic distance effect was a significant predictor of fluency, $\beta = -0.236$, $p<0.05$, whereas the nonsymbolic distance effect did not predict fluency, $\beta=0.063$, $p>0.05$.

@@

In our final example, we demonstrate a Bayesian factorial analysis of variance. To do this, we use a data set based on @campbellFugelsang2001, who investigated the effects of problem size and numerical surface form on response times in an addition verification task. Not only did they find the usual effects of problem size (small problems are solved faster than large problems) and problem format (digit problems are solved faster than word problems), but they also found an interaction between size and format; specifically, the problem size effect was larger for problems in word format than for problems in digit format. This interaction was interpreted as support for an *interactive* model of mental arithmetic, where factors that influence encoding processes (e.g., surface format) also directly influence calculation processes.

@@

The examples provided in this tutorial represent a broad class of techniques that are common to many studies in numerical cognition and should be familiar to most researchers.

- reiterate benefits of Bayesian testing
- give reference to other JASP and Bayes tutorials

### 2019-12-07 jncTutorial/manuscript.tex
minor edits throughout and start conclusion

Like any other discipline within the psychological and behavioral sciences, researchers in numerical cognition rely on statistical inference, and hypothesis testing in particular, to verify theoretical claims against observed data. In recent years, Bayesian inference has become increasingly popular as an alternative to classical null hypothesis significance testing (NHST). Bayesian inference operates under a fundamentally different framework from NHST; instead of hard \enquote{accept/reject} decisions about a null hypothesis, Bayesian inference works by quantifying knowledge about competing hypotheses in light of observed data. Hypotheses that do a better job of predicting data will receive increased support, whereas hypotheses that don't predict the data well will receive decreased support. Under this framework, researchers can mathematically quantify belief about parameters

@@

Even though the reader of this paper is probably familiar with the general concept of hypothesis testing, we believe that it is nonetheless instructive to go over the basics.

@@

Hypothesis testing can be viewed as a process for validating general claims based on a small sample of the general population. For illustrative purposes, suppose that we are policy makers confronted with a new study programme which claims to increase the mathematical abilities of sixth graders in the United States. To ease the exposition we use a (fictitious) scale called the Scale for Advancing Mathematical Ability (SAMA) to measure mathematical ability. As policy makers we know that the population mean SAMA score is \(\mu_{0}=50\) and that the standard deviation is \(\sigma=10\) for the population of sixth graders that are taught using the standard materials in the US.

@@

To nonetheless infer the validity of the working hypothesis \(\mathcal{H}_{1}\), we run an experiment in which \(n=90\) sixth graders are taught with the new study programme. Suppose that these particular participants scored an average SAMA score of \(\bar{x} = 52.07\). Clearly, this sample mean is larger than the status quo population mean of \(50\), but can we assume that this finding generalises to the whole population? If the answer is \enquote{yes} and we declare the working hypothesis to be verified, then how confident would we be about such a decision? Below we recall how \(p\)-values are typically used to make hard \enquote{yes} and \enquote{no} decisions, and we remind the reader why a \(p\)-value does not quantify the (un)certainty of such decisions.

@@

To deduce how hypotheses, i.e., claims about the population such as \(\mu\), affect the data, such as \(\bar{x}\), we have to link the former to the latter with a distribution, for instance, with a normal distribution. The population parameters, the distribution that links them to the data, and the data together make up a (statistical) model, which allows for quantitative predictions. For instance, the claim that the treatment population is identitical to the current population of sixth graders implies the null model \(\mathcal{M}_{0}\) comprised of (1) the null hypothesis \(\mathcal{H}_{0}: \mu = 50\), (2) the assumption \(\sigma = 10\), and (3) normally distributed SAMA scores. When all three assumptions hold true, we then deduce that the \emph{potential} average SAMA score consisting of a group of \(n = 90\) participants is expected to fall within the interval \(47.93 = 50 - 1.96 \frac{10}{\sqrt{n}}\) and \(52.07 = 50 + 1.96 \frac{10}{\sqrt{n}}\) with 95\% chance.
\footnote{We use ``chance'' to refer to probabilistic statements about the data.}

@@

Reasoning from the general population under the assumption that the parameters are known to derive statements about potential outcomes forms the basis of null hypothesis signficance testing. By definition the \(p\)-value is the chance of seeing the observed outcome and more extreme --but \emph{unobserved}-- outcomes under the null.

@@

Bayesian testing addresses these problems as follows: Firstly, it compares the two hypotheses on equal footing by also operationalising the alternative model. For the case at hand, \(\mathcal{M}_{1}\) would be comprised of the assumptions (1') \(\mathcal{H}_{1}: \mu > 50\), (2) \(\sigma =10\) as before, and (3) the data being normally distributed as before. By doing so, violations of normality and the assumed standard deviation will penalize both models equally. Secondly, to reason from a particular observed outcome to the general population, it employs Bayes rule. To do so, we are required to express our uncertainty about the two hypotheses in terms of so-called prior model probabilities before data observation.

Prior model probabilities help us contextualise the testing problem. Working as policy makers, we might have seen many claims of improved study programmes that did not work out well. For instance, if 4-to-1 claims of improved study programmes led to null results, we might set \(P( \mathcal{M}_{1}) = 0.2\) and \(P(\mathcal{M}_{0}) = 0.8\). On the other hand, the designers of the new study programme to increase SAMA scores might choose \(P( \mathcal{M}_{1}) = 0.7\) and \(P(\mathcal{M}_{0}) = 0.3\) instead. If \(\mathcal{M}_{1}\) and \(\mathcal{M}_{0}\) are the only models under consideration we have \(0 < P( \mathcal{M}_{0}) < 1\) and \(P( \mathcal{M}_{1})= 1 - P( \mathcal{M}_{0})\). The Bayes factor \(\textnormal{BF}_{10}(d)\) is a function of the observed data \(d\) that updates these prior model probabilities to posterior model probabilities, as follows:

@@

Prior model probabilities are context dependent, whereas Bayes factors only rely on the statistical design. To simplify notation we drop the reference to the data in the Bayes factor and denote them as \(\textnormal{BF}_{10}\), where the 1 in the subscript refers to the alternative model, and the 0 to the null model. Bayes factor ranges from zero to \(+ \infty\), and \(\textnormal{BF}_{10} = 8\) implies that the observed data are 8 times more likely under the alternative compared to the null model. For Bayes factors smaller than one, we typically take the reciprocal to make them easier to interpret. For instance, \(\textnormal{BF}_{10} = 0.2\) is easier interpret as \(\textnormal{BF}_{01} = 1/\textnormal{BF}_{10} = 5\), which means that the observed data are 5 times more likely under the null compared to the alternative. Bayes factors are best viewed as a continuous measure of evidence. In general, a value of \(\textnormal{BF}_{10}\) larger than one is perceived to provide evidence for the alternative against the null. Also, if \(\textnormal{BF}_{10}\) grows to \(\infty\), the posterior model probability \(P(\mathcal{M}_{1} \mid d)\) grows to one, thus the more certain we will be about the adequacy of the alternative.

@@

What is important to realize is that a test addresses a different question than an estimation. The Bayes factor focuses on the question \enquote{Does the effect exists?}, whereas the posterior focuses on the question \enquote{If it exists, then how large is it?}. This is why the (Bayesian) testing problem compares the model where the population effect is fixed at zero against one where it is free to vary. Once the data have convinced us that the population effect \(\delta\) varies freely would we be interested in the magnitude of the effect.

@@

We use an example inspired by Holloway and Ansari (2009), who examined the relationship between basic numerical processing and higher mathematical skills in children. Specifically, they examined whether childrens' mathematical achievement scores was related to their ability to compare symbolic and nonsymbolic number magnitudes. In this example, we will recreate one specific component of their analysis; namely, we will use Bayesian linear regression to examine the extent to which mathematical fluency scores are predicted by childrens' standardized symbolic and nonsymbolic distance effects, calculated by subtracting mean RTs for large numerical distances from mean RTs for small numerical distances and then dividing by the mean RT for large distances. In their original paper, Holloway and Ansari (2009) found that the childrens' symbolic distance effect was a significant predictor of fluency, \(\beta = -0.236\), \(p<0.05\), whereas the nonsymbolic distance effect did not predict fluency, \(\beta=0.063\), \(p>0.05\).

@@

To demonstrate a Bayesian factorial analysis of variance, we use a data set based on Campbell and Fugelsang (2001), who investigated the effects of problem size and numerical surface form on response times in an addition verification task. In addition to the usual effects of problem size (small problems are solved faster than large problems) and problem format (digit problems are solved faster than word problems), they found an interaction between size and format, which they interpreted as support for an interactive model of mental arithmetic where factors that affect encoding processes (e.g., surface format) can directly impact calculation processes.

@@

>>>>

Like any other discipline within the psychological and behavioral sciences, researchers in numerical cognition rely on statistical inference, and hypothesis testing in particular, to verify theoretical claims against observed data. In recent years, Bayesian inference has become increasingly popular as an alternative to classical null hypothesis significance testing (NHST). Bayesian inference operates under a fundamentally different framework from NHST; instead of hard \enquote{accept/reject} decisions about a null hypothesis, Bayesian inference works by quantifying knowledge about competing hypotheses in light of observed data. Hypotheses that do a better job of predicting data will receive increased support, whereas hypotheses that don't predict the data well will receive decreased support.

@@

In this section, we will review and elaborate on the concept of hypothesis testing. Even though the reader of this paper is probably familiar with the general concept of hypothesis testing, we believe that it is nonetheless instructive to go over the basics.

@@

Hypothesis testing can be viewed as a process for validating general claims based on a small sample of the general population. To illustrate, suppose that we are policy makers confronted with a new study programme which claims to increase the mathematical abilities of sixth graders in the United States. To ease the exposition we use a (fictitious) scale called the Scale for Advancing Mathematical Ability (SAMA) to measure mathematical ability. As policy makers we know that the population mean SAMA score is \(\mu_{0}=50\) and that the standard deviation is \(\sigma=10\) for the population of sixth graders that are taught using the standard materials in the US.

@@

Nonetheless, we can infer the validity of the working hypothesis \(\mathcal{H}_{1}\) by running an experiment in which \(n=90\) sixth graders are taught with the new study programme. Suppose that these particular participants scored an average SAMA score of \(\bar{x} = 52.07\). Clearly, this sample mean is larger than the status quo population mean of \(50\), but can we assume that this finding generalises to the whole population? If the answer is \enquote{yes} and we declare the working hypothesis to be verified, then how confident would we be about such a decision? Below we recall how \(p\)-values are typically used to make hard \enquote{yes} and \enquote{no} decisions, and we remind the reader why a \(p\)-value does not quantify the (un)certainty of such decisions.

@@

To deduce how hypotheses, i.e., claims about the population such as \(\mu\), affect the data, such as \(\bar{x}\), we have to link the former to the latter with a distribution. For instance, let us suppose a normal distribution provides this link. The population parameters, the distribution that links them to the data, and the data together make up a (statistical) model, which allows for quantitative predictions. For instance, the claim that the treatment population is identitical to the current population of sixth graders implies the null model \(\mathcal{M}_{0}\) comprised of (1) the null hypothesis \(\mathcal{H}_{0}: \mu = 50\), (2) the assumption \(\sigma = 10\), and (3) normally distributed SAMA scores. When all three assumptions hold true, we can mathematically deduce that the \emph{potential} average SAMA score consisting of a group of \(n = 90\) participants is expected to fall within the interval \(47.93 = 50 - 1.96 \frac{10}{\sqrt{n}}\) and \(52.07 = 50 + 1.96 \frac{10}{\sqrt{n}}\) with 95\% chance.
\footnote{We use ``chance'' to refer to probabilistic statements about the data.}

@@

Reasoning from the general population under the assumption that the parameters are known to derive statements about potential outcomes forms the basis of null hypothesis signficance testing. By definition, the \(p\)-value is the chance of seeing the observed outcome and more extreme --but \emph{unobserved}-- outcomes under the null.

@@

Bayesian testing addresses these problems as follows. First, it compares the two hypotheses on equal footing by also operationalising the alternative model. For the case at hand, \(\mathcal{M}_{1}\) would be comprised of the assumptions (1') \(\mathcal{H}_{1}: \mu > 50\), (2) \(\sigma =10\) as before, and (3) the data being normally distributed as before. By doing so, violations of normality and the assumed standard deviation will penalize both models equally. Secondly, to reason from a particular observed outcome to the general population, it employs Bayes rule. To do so, we are required to express our uncertainty about the two hypotheses in terms of so-called \emph{prior model probabilities} before data observation.

Prior model probabilities help us contextualise the testing problem. Working as policy makers, we might have seen many claims of improved study programmes that did not work out well. For instance, past claims of improved study programmes led to null results four times out of five (i.e., we place 4-to-1 odds against the claim of improvement), we might set \(P( \mathcal{M}_{1}) = 0.2\) and \(P(\mathcal{M}_{0}) = 0.8\). On the other hand, the designers of the new study programme to increase SAMA scores might feel strongly about the benefits of the problem, and so might choose \(P( \mathcal{M}_{1}) = 0.7\) and \(P(\mathcal{M}_{0}) = 0.3\) instead. If \(\mathcal{M}_{1}\) and \(\mathcal{M}_{0}\) are the only models under consideration we have \(0 < P( \mathcal{M}_{0}) < 1\) and \(P( \mathcal{M}_{1})= 1 - P( \mathcal{M}_{0})\). The Bayes factor \(\textnormal{BF}_{10}(d)\) is a function of the observed data \(d\) that updates these prior model probabilities to posterior model probabilities, as follows:

@@

Prior model probabilities are context dependent, whereas Bayes factors only rely on the statistical design. To simplify notation we drop the reference to the data in the Bayes factor and denote it as \(\textnormal{BF}_{10}\), where the 1 in the subscript refers to the alternative model, and the 0 to the null model. Bayes factors range from zero to \(+ \infty\), and \(\textnormal{BF}_{10} = 8\) implies that the observed data are 8 times more likely under the alternative compared to the null model. For Bayes factors smaller than one, we typically take the reciprocal to make them easier to interpret. For instance, \(\textnormal{BF}_{10} = 0.2\) is easier interpret as \(\textnormal{BF}_{01} = 1/\textnormal{BF}_{10} = 5\), which means that the observed data are 5 times more likely under the null compared to the alternative. Bayes factors are best viewed as a continuous measure of evidence. In general, a value of \(\textnormal{BF}_{10}\) larger than one is perceived to provide evidence for the alternative against the null. Also, if \(\textnormal{BF}_{10}\) grows to \(\infty\), the posterior model probability \(P(\mathcal{M}_{1} \mid d)\) grows to one, thus the more certain we will be about the adequacy of the alternative.

@@

What is important to realize is that a test addresses a different question than an estimation. The Bayes factor focuses on the question \enquote{Does the effect exist?}, whereas the posterior focuses on the question \enquote{If it exists, then how large is it?}. This is why the (Bayesian) testing problem compares the model where the population effect is fixed at zero against one where it is free to vary. Once the data have convinced us that the population effect \(\delta\) varies freely, only then would we be interested in the magnitude of the effect.

@@

We use an example inspired by Holloway and Ansari (2009), who examined the relationship between basic numerical processing and higher mathematical skills in children. Specifically, they examined whether childrens' mathematical achievement scores were related to their ability to compare symbolic and nonsymbolic number magnitudes. In this example, we will recreate one specific component of their analysis; namely, we will use Bayesian linear regression to examine the extent to which mathematical fluency scores are predicted by childrens' standardized symbolic and nonsymbolic distance effects, calculated by subtracting mean RTs for large numerical distances from mean RTs for small numerical distances and then dividing by the mean RT for large distances. In their original paper, Holloway and Ansari (2009) found that the childrens' symbolic distance effect was a significant predictor of fluency, \(\beta = -0.236\), \(p<0.05\), whereas the nonsymbolic distance effect did not predict fluency, \(\beta=0.063\), \(p>0.05\).

@@

In our final example, we demonstrate a Bayesian factorial analysis of variance. To do this, we use a data set based on Campbell and Fugelsang (2001), who investigated the effects of problem size and numerical surface form on response times in an addition verification task. Not only did they find the usual effects of problem size (small problems are solved faster than large problems) and problem format (digit problems are solved faster than word problems), but they also found an interaction between size and format; specifically, the problem size effect was larger for problems in word format than for problems in digit format. This interaction was interpreted as support for an \emph{interactive} model of mental arithmetic, where factors that influence encoding processes (e.g., surface format) also directly influence calculation processes.

@@

The examples provided in this tutorial represent a broad class of techniques that are common to many studies in numerical cognition and should be familiar to most researchers.

\begin{itemize}
\tightlist
\item
  reiterate benefits of Bayesian testing
\item
  give reference to other JASP and Bayes tutorials
\end{itemize}
